Practical Optional Types for Clojure

### 2015-02-27 overview.tex
overview

The next example is extracted from a production system at CircleCI
(section~\ref{sec:casestudy} presents a case study).

@@

Next, 
HMaps interact with occurrence typing in interesting ways.

\begin{exmp}
\inputminted[firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}

Section~\ref{sec:hmapformal} discusses this in greater detail.

@@

The flexibility of \clj{isa?} is key to the generality of multimethods. 
We can dispatch on the \clj{:op} key 
of our HMap AST \clj{Expr}.
Since keywords are functions that look themselves up in their argument, we simply
use \clj{:op} as the dispatch function.

\begin{exmp}
\inputminted[firstline=5,lastline=31]{clojure}{code/demo/src/demo/eg5.clj}
\end{exmp}

\clj{isa?} is special with vectors---vectors of the
same length recursively call \clj{isa?} on the elements pairwise.
\begin{minted}{clojure}
  (isa? [Keyword Keyword] [Object Object]) ;=> true
\end{minted}

Now we can simulate multiple dispatch by dispatching on
a vector containing the class of both arguments.

@@

We call three different \clj{File} constructors, each known at compile-time
via type hints.
No extra annotations are needed to follow type-directed control flow
in multimethod dispatch.

>>>>

\paragraph{HMaps in Practice} The next example is extracted from a production system at CircleCI
(section~\ref{sec:casestudy} presents a case study).

@@

We would still catch the bug if \clj{dissoc} forgot to remove \clj{:private-key},
but now stashing the raw private key as \clj{:secret-key} is allowed.

\paragraph{Branching on HMaps} Finally, testing on HMap properties
allows us to refine its type down branches. \clj{dec-map} takes an
\clj{Expr}, traverses to its nodes and decrements their values, then
builds the \clj{Expr} back up with the decremented nodes.

\begin{exmp}
\inputminted[linenos,firstnumber=1,firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
\label{example:decmap}
\end{exmp}

If we go down the then branch (line 4), since \clj{(= (:op m) :if)} is true
we remove
the \clj{:do} and \clj{:const}
Expr's from the type of \clj{m} (because their respective \clj{:op} entries disagrees with \clj{(= (:op m) :if)})
and we are left with an \clj{:if} Expr.
On line 8,
we instead strike out the \clj{:if} Expr since it contradicts \clj{(= (:op m) :if)} being false. 
Line 9 we know \clj{m}
we remove \clj{:const} because it contradicts \clj{(= (:op m) :do)} being true,
and we know \clj{m} is a \clj{:do} Expr.
Line 12 we strike out \clj{:do} because \clj{(= (:op m) :do)} is false,
so we are left with \clj{m} being a \clj{:const} Expr.

Section~\ref{sec:formalpaths} discusses how this automatic reasoning is achieved.

@@

\paragraph{HMap dispatch} The flexibility of \clj{isa?} is key to the generality of multimethods. 
We can dispatch on the \clj{:op} key 
of our HMap AST \clj{Expr}.
Since keywords are functions that look themselves up in their argument, we simply
use \clj{:op} as the dispatch function.

\begin{exmp}
\inputminted[firstline=5,lastline=18]{clojure}{code/demo/src/demo/eg5.clj}
\label{example:incmap}
\end{exmp}

\clj{inc-map} is like example~\ref{example:decmap} except the nodes are incremented.
The reasoning is similar, except we only use positive
information since we only consider the \emph{dispatch value} and how it relates
to the \emph{dispatch function}.
In the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}
and \clj{:const} methods.

\paragraph{Multiple dispatch} \clj{isa?} is special with vectors---vectors of the
same length recursively call \clj{isa?} on the elements pairwise.
\begin{minted}{clojure}
  (isa? [Keyword Keyword] [Object Object]) ;=> true
\end{minted}

Now we can simulate multiple dispatch by dispatching on
a vector containing the class of both arguments. \clj{open}
takes two arguments which can be strings or files and returns
a new file that concatenates their paths.

@@

We call three different \clj{File} constructors, each known at compile-time
via type hints.
Multiple dispatch follows the same kind of reasoning as example~\ref{example:incmap},
except we update multiple bindings simultaneously.
